The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-08-07 paper/referee-reply.tex

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{graphicx}

\title{Response to the second report on the article draft \emph{The
    Impact of Gas Bulk Rotation on the Lyman-$\alpha$ line.}} 
\author{Juan N. Garavito-Camargo, Jaime E. Forero-Romero, Mark Disjkstra}
\date{\today}

@@

In what follows the comments by the referee are boldfaced.

Best regards, \\

The Authors\\

\section*{Global Comments}

\subsection*{Variations with viewing angle... again}

{\bf Figs 2 and 3 are very nice. They illustrate clearly the evolution of spectral shape with viewing angle. Note that the x label is wrong, it is not Vmax. It also needs a color bar to explicit the scale between a red pixel and a green/blue pixel. Because... to my eyes, there is an obvious third result illustrated on these Figs, which is that much less photons seem to escape at the equator than along the poles: the red pixels are concentrated towards the pole, whereas there seems to be less light emerging from equatorial directions. But if the dynamical range between red and blue is very small, maybe it’s ok, given that the equatorial spectrum is broader...
How did you select photons when you did this Fig 2 ? Did you select them on $\cos\theta|$? You should select only one side, one hemisphere, otherwise you count twice more photons anywhere else than along the equator. Did you select a peculiar $\varphi$ ? Or did you sum over the azimuthal direction ? In order to avoid this impression that less photons escape at equator, you should indeed sum over $\varphi$. However, these selection biases would also affect the static case, whereas the first column looks ok to me, it seems that the same “intensity” is escaping from all directions for this column, and not for the others.
It is in strong contradiction with your Fig 5. How do you explain this ?
Furthermore, I would have expected the contrary, with more photons escaping at the equator, where the velocity field is stronger, than through the poles, where everything is static... However, Ly$\alpha$ radiation transfer effects can often be far from intuitiv.}

\subsection*{intrinsic spectrum on Fig 4}

{\bf From your Figs 6 and 7, it seems that the results for central and homogeneous cases are very comparable. I would like to see the shape and FWHM of the intrinsic spectra (I mean before transfer) in the case of homogeneous sources, for the static case, as the fiducial case, and when Vmax=100,200,300 km/s, along the equatorial direction. I would like to infer how much of the broadening of the line is really due to transfer effects. Could you make a figure where you compare the intrinsic and emerging spectra along
the equatorial plane ?}

\subsection*{Sect 4.1 : Rotation = Static ?}

{\bf I don’t understand the reasonning that rotation would act on the Ly$\alpha$ transfer as a static medium, since Ly$\alpha$ photons are “co-rotating”. You mention this argument once in the “answer to referee” document, and once in the text Sect. 4.1, with a sentence which reads absurd : a rotating sphere is identical from a static one. At least, it is not well formulated.
It seems to me that your argument should be right for any kind of motion. Why would it be specific to rotation ? In other words, you could exchange the word “rotation” with the word “outflow” or “inflow” in your reasonning, right ? “While scattering events off atoms within the outflowing cloud impart Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only there in the lab-frame. Therefore, in the frame of the outflowing gas cloud all atoms are stationnary with respect to each other and the scattering process proceeds identical as in the static case.”. However, all atoms in a rotating/outflowing/infalling cloud are not stationary with respect to each other, otherwise the cloud would be static! Your reasonning is true at the scale of each cell, but not at global scale.
I did not understand the alternative -more quantitative- explanation better.
Can you develop this point ? Maybe, to convince me, you could look at the redistribution of frequencies after one single diffusion ( 1/ in all directions, 2/ along the rotation axis, 3/ in the equator plane), emitting your photons at a radius r = 0.9 R, and comapre them with the same redistributions in the static case. If they are identical, you are right, rotation does not have any effect... if they are different, multiple scatterings will have a cumulative effect on the spectral shape, but also on the probalility to escape more easily in some directions than others... I understand that this implies new simulations and more time, but I would really appreciate to see these plots. }

\subsection*{Angular variation of the distribution of escaping directions}

{\bf It seems to me that Ly$\alpha$ photons escape a medium through the path of minimum optical depth. So, in a rotating cloud, where the velocity field is tangential, they might “rotate” or spiral from the center to the edge. To test this idea, you could look at the distribution of escaping directions from photons emerging at the pole or at the equator. But for this, you need to keep track of the location of escape, and not only of the direction of escape. Did you collect this information ?
I would expect that photons at the equator escape more tangentially to the sphere than at the poles... But again, Ly$\alpha$ transfer is not exactly intuitiv.}

@@

\section*{Details}

\subsection*{Sect. 3.1}

{\bf You say that “If the viewing angle is aligned with the rotation axis, $|\mu| \approx 1$, the Ly$\alpha$ line keeps in most of the cases a double peak”, isn’t it in all cases ?}

\subsection*{Sect. 3.5}
{\bf The parameter a is not defined when you discuss Neufeld.}

\subsection*{Fig9}

{\bf Is Fig 9, left panel, for a central source ? Can you show the homogeneous case also ? You discuss in the referee repport that the distribution is not bimodal anymore. But it has to be different than in central case, spanning the whole range of nb scatt, because photons are emitted at every radii in the cloud, so at every optical depth.
On the right panel, solid and dashed lines are not defined.}

\subsection*{Vmax = 1000 km/s}
{\bf If you did this test and checked that the nb of scatterings and the escape fraction is still angle invariant for such a high velocity, then it is worth mentionning it. Because in principle, media with very high velocities should start to become transparent for Ly$\alpha$.}

### 2014-08-07 paper/referee-reply.tex

The Authors\\

\section*{Global Comments}

\subsection*{Variations with viewing angle... again}

{\bf Figs 2 and 3 are very nice. They illustrate clearly the evolution
  of spectral shape with viewing angle. Note that the x label is
  wrong, it is not Vmax. It also needs a color bar to explicit the
  scale between a red pixel and a green/blue pixel. Because... to my
  eyes, there is an obvious third result illustrated on these Figs,
  which is that much less photons seem to escape at the equator than
  along the poles: the red pixels are concentrated towards the pole,
  whereas there seems to be less light emerging from equatorial
  directions. But if the dynamical range between red and blue is very
  small, maybe it’s ok, given that the equatorial spectrum is
  broader}

We have correctd the xlabel. Indeed the
concentration of photons in a narrow range of velocities along the
poles (i.e. red values) is compensated by a broader distribution
around the equator. The total number of photons is the same regardless
of the $\cos\theta$ value. This is the result of the test shown in
Figure 5.

{\bf How did you select photons when you did this Fig 2 ? Did you
  select them on $|\cos\theta|$? You should select only one side, one
  hemisphere, otherwise you count twice more photons anywhere else
  than along the equator. Did you select a peculiar $\varphi$ ? Or did
  you sum over the azimuthal direction ? In order to avoid this
  impression that less photons escape at equator, you should indeed
  sum over $\varphi$. }

The 2D distribution is symmetric on $\cos\theta$. We mention this in
the first paragraph of the results section. Using this symmetry we select
by $|\cos\theta|$ in order to have less noise in the 2D histograms. We
also sum over $\varphi$.

{\bf However, these selection biases would also affect the static case, whereas the first column looks ok to me, it seems that the same “intensity” is escaping from all directions for this column, and not for the others.
It is in strong contradiction with your Fig 5. How do you explain this ?
Furthermore, I would have expected the contrary, with more photons escaping at the equator, where the velocity field is stronger, than through the poles, where everything is static... However, Ly$\alpha$ radiation transfer effects can often be far from intuitiv.}

The colors in the Figures 2 and 3 correspond to number densities
inside a velocity-$|\cos\theta|$ pixel. The total flux is thus an
integrated quantity over the velocity and should not be estimated from
the pixel intensity. It is difficult to visualize
the result from Figures 2 and 3. That's why we prepare Figure 5 to
show that the integrated flux is independent of the angle.

@@

\subsection*{intrinsic spectrum on Fig 4}

{\bf From your Figs 6 and 7, it seems that the results for central and homogeneous cases are very comparable. I would like to see the shape and FWHM of the intrinsic spectra (I mean before transfer) in the case of homogeneous sources, for the static case, as the fiducial case, and when Vmax=100,200,300 km/s, along the equatorial direction. I would like to infer how much of the broadening of the line is really due to transfer effects. Could you make a figure where you compare the intrinsic and emerging spectra along
the equatorial plane ?}

The Figure is here:\\

\includegraphics[width=1.0\textwidth]{../code/intrinsic_spectrum.png}

### 2014-08-08 paper/referee-reply.tex

{\bf From your Figs 6 and 7, it seems that the results for central and
  homogeneous cases are very comparable. I would like to see the shape
  and FWHM of the intrinsic spectra (I mean before transfer) in the
  case of homogeneous sources, for the static case, as the fiducial
  case, and when Vmax=100,200,300 km/s, along the equatorial
  direction. I would like to infer how much of the broadening of the
  line is really due to transfer effects. Could you make a figure
  where you compare the intrinsic and emerging spectra along the
  equatorial plane ?}

The Figure is here:\\

\includegraphics[width=1.0\textwidth]{../code/intrinsic_spectrum.png}

Each column corresponds to $100$, $200$ and $300$km/s. The intrinsic
spectrum for the static case is a delta function around $x=0$.

@@

\subsection*{Sect 4.1 : Rotation = Static ?}

{\bf I don’t understand the reasonning that rotation would act on the
  Ly$\alpha$ transfer as a static medium, since Ly$\alpha$ photons are
  “co-rotating”. You mention this argument once in the “answer to
  referee” document, and once in the text Sect. 4.1, with a sentence
  which reads absurd : a rotating sphere is identical from a static
  one. At least, it is not well formulated.

It seems to me that your argument should be right for any kind of
motion. Why would it be specific to rotation ? In other words, you
could exchange the word “rotation” with the word “outflow” or “inflow”
in your reasonning, right ? “While scattering events off atoms within
the outflowing cloud impart Doppler boosts on the Ly$\alpha$ photon,
these Doppler boost are only there in the lab-frame. Therefore, in the
frame of the outflowing gas cloud all atoms are stationnary with
respect to each other and the scattering process proceeds identical as
in the static case.”. However, all atoms in a
rotating/outflowing/infalling cloud are not stationary with respect to
each other, otherwise the cloud would be static! Your reasonning is
true at the scale of each cell, but not at global scale.

I did not understand the alternative -more quantitative- explanation better.
Can you develop this point ? Maybe, to convince me, you could look at
the redistribution of frequencies after one single diffusion ( 1/ in
all directions, 2/ along the rotation axis, 3/ in the equator plane),
emitting your photons at a radius r = 0.9 R, and comapre them with the
same redistributions in the static case. If they are identical, you
are right, rotation does not have any effect... if they are different,
multiple scatterings will have a cumulative effect on the spectral
shape, but also on the probalility to escape more easily in some
directions than others... I understand that this implies new
simulations and more time, but I would really appreciate to see these
plots. } \\

These are good points. The argument that you can replace the word
'rotating' with 'outflowing' was interesting, and briefly confused
us. However, there is a clear difference between the two cases. For a
cloud undergoing solid body rotation, we can draw a line between any
two atoms within it, and their relative velocity along this line is
zero (apart from the relative velocity as a result of Thermal motion),
irrespective of the rotation velocity of the cloud. This relative
velocity is what is relevant for the RT. In contrast, if one repeats
the same exercise for the outflowing/inflowing case, then will be a
relative velocity between two atoms which depends on the
outflow/inflow velocity.\\

We can further illustrate this point by having one photon be emitted
in the center of the rotating cloud into the plane of the equator, and
another in the direction of the pole. Let both photons travel $\tau=1$
(at line center), and then scatter by 90 degrees, preserve its
frequency in the frame of the atom, and then again travel $tau=1$. In
both cases, the photon travels the same distance. In the LAB-frame,
both photons will have different frequencies, but not in the frame of
the gas.\\

The situation described above is clearly unique to solid-body
rotation. It does not apply to non-solid body rotation, which we
intend to study in upcoming work. We think our own + the referees
confusion only emphasizes that even this `simplified' problem of
solid-body rotation is highly non-trivial. We have attempted to
clarify this further in \S 3.X

@@

\subsection*{Angular variation of the distribution of escaping directions}

{\bf It seems to me that Ly$\alpha$ photons escape a medium through
  the path of minimum optical depth. So, in a rotating cloud, where
  the velocity field is tangential, they might “rotate” or spiral from
  the center to the edge. To test this idea, you could look at the
  distribution of escaping directions from photons emerging at the
  pole or at the equator. But for this, you need to keep track of the
  location of escape, and not only of the direction of escape. Did you
  collect this information ?

I would expect that photons at the equator escape more tangentially to
the sphere than at the poles... But again, Ly$\alpha$ transfer is not
exactly intuitiv.}

We have looked into this. We selected two different sets of
photons. The first one escapes through the north pole and the second
around the equator in a region with $0<\varphi<90$. For each set we
compute the dot product $\mu\equiv\hat{k}_{out}\cdot\hat{r}_{out}$,
where $\hat{r}_{out}$ is the unit vector in the radial direction for
each photon. We plot the distributions for $\mu$ for two different
$\tau_H$. In each plot we find that the two distributions (equator and
pole) are virtually the same:

\begin{center}
\includegraphics[width=0.6\textwidth]{../code/surface_mu_tau5_diff_pos.png}
\includegraphics[width=0.6\textwidth]{../code/surface_mu_tau7_diff_pos.png}
\end{center}

The distribution is also insensitive to the rotational velocity:

\begin{center}
\includegraphics[width=0.6\textwidth]{../code/surface_mu_t5_diff_v.png}
\end{center}

The only trend can be found for different optical depths:

\begin{center}
\includegraphics[width=0.6\textwidth]{../code/surface_mu_diff_tau.png}
\end{center}

@@

\subsection*{Fig9}

{\bf Is Fig 9, left panel, for a central source ? Can you show the
  homogeneous case also ? You discuss in the referee repport that the
  distribution is not bimodal anymore. But it has to be different than
  in central case, spanning the whole range of nb scatt, because
  photons are emitted at every radii in the cloud, so at every optical
  depth. On the right panel, solid and dashed lines are not defined.}

\subsection*{Vmax = 1000 km/s}
{\bf If you did this test and checked that the nb of scatterings and
  the escape fraction is still angle invariant for such a high
  velocity, then it is worth mentionning it. Because in principle,
  media with very high velocities should start to become transparent
  for Ly$\alpha$.}

For solid body rotation, the results should not depend on $v_{\rm
  max}$ at all. We have even done calculations with $v_{\rm max}=10^6$
km/s, and confirmed this result.

### 2014-08-08 paper/referee-reply.tex

\subsection*{intrinsic spectrum on Fig 4}

{\bf From your Figs 6 and 7, it seems that the results for central and
  homogeneous cases are very comparable. I would like to see the shape
  and FWHM of the intrinsic spectra (I mean before transfer) in the
  case of homogeneous sources, for the static case, as the fiducial
  case, and when Vmax=100,200,300 km/s, along the equatorial
  direction. I would like to infer how much of the broadening of the
  line is really due to transfer effects. Could you make a figure
  where you compare the intrinsic and emerging spectra along the
  equatorial plane ?}

The Figure is here:\\

\includegraphics[width=1.0\textwidth]{./intrinsic_spectrum.png}

@@

{\bf It seems to me that Ly$\alpha$ photons escape a medium through
  the path of minimum optical depth. So, in a rotating cloud, where
  the velocity field is tangential, they might “rotate” or spiral from
  the center to the edge. To test this idea, you could look at the
  distribution of escaping directions from photons emerging at the
  pole or at the equator. But for this, you need to keep track of the
  location of escape, and not only of the direction of escape. Did you
  collect this information ?

I would expect that photons at the equator escape more tangentially to
the sphere than at the poles... But again, Ly$\alpha$ transfer is not
exactly intuitiv.}

We have looked into this. We selected two different sets of
photons. The first one escapes through the north pole and the second
around the equator in a region with $0<\varphi<90$. For each set we
compute the dot product $\mu\equiv\hat{k}_{out}\cdot\hat{r}_{out}$,
where $\hat{r}_{out}$ is the unit vector in the radial direction for
each photon. We plot the distributions for $\mu$ for two different
$\tau_H$. In each plot we find that the two distributions (equator and
pole) are virtually the same:

\begin{center}
\includegraphics[width=0.6\textwidth]{./surface_mu_tau5_diff_pos.png}
\includegraphics[width=0.6\textwidth]{./surface_mu_tau7_diff_pos.png}
\end{center}

@@

The distribution is also insensitive to the rotational velocity:

\begin{center}
\includegraphics[width=0.6\textwidth]{./surface_mu_t5_diff_v.png}
\end{center}

@@

The only trend can be found for different optical depths:

\begin{center}
\includegraphics[width=0.6\textwidth]{./surface_mu_diff_tau.png}
\end{center}

### 2014-08-11 paper/ms.tex
minor change in reference
minor updates in ms
Mark's input

A basic understanding of the quantitative behavior of the \ly line
has been reached through analytic studies in the case of a static
configurations, such as uniform slabs
\citep{Adams72,Harrington73,Neufeld90} and uniform spheres
\citep{Dijkstra06}. 
Analytic studies of configurations including some kind of bulk flow
only include the case of a sphere with a Hubble like expansion flow
\citep{LoebRybicki}.

A more detailed quantitative description of the \ly line has been
reached through Monte Carlo simulations \citep{Auer68,Avery68,Adams72}. 
In the last two decades these studies have become popular due to the
availability of computing power. 
Early into the 21st century, the first
studies focused on homogeneous and static media
\citep{Ahn00,Ahn01,Zheng02}. 
Later on, the effects of clumpy media \citep{Hansen06} and
expanding/contracting shell/spherical geometries started to be studied
\citep{Ahn03,Verhamme06,Dijkstra06}.  
Similar codes have applied these results to semi-analytic models of
galaxy formation \citep{Orsi12, Garel2012} and results of large
hydrodynamic simulations \citep{CLARA,Forero12,Behrens13}. 
Recently, Monte Carlo codes have also been applied to the results of
high resolution hydrodynamic simulations of individual galaxies
\citep{Laursen09,Barnes11,Verhamme12,Yajima12}. 
Meanwhile, recent  developments have been focused on the systematic
study of clumpy outflows \citep{DijkstraKramer} and anisotropic
velocity configurations \citep{Zheng2013}.

@@

\subsection{Integrated Line Intensity}

@@

Fig. \ref{fig:muhisto} shows the results for a selection of models
with $\tau_{\rm H}=10^{5}$, different rotational velocities and the two
types of source distributions.
This shows that $F(\mu)$ is consistent with being flat, apart
from some statistical fluctuations on the order of \%2.

This is a remarkable result. 
In spite of that the rotation axis defines preferential direction, the
integrated flux is the same for all viewing angles in the
range of parameters explored in this paper.

@@

We now estimate the escape fraction $f_{\rm esc}$ for the dusty
models. 
The main  result is that we do not find any significant dependence
with either the viewing angle nor the rotational velocity.  
For completeness we list in Table \ref{table:escape} the scape
fraction for all models.

@@

The number of scatterings affects the escape frequency of a \ly
photon. 
Studying this quantity can help us in clarifying the independence of
the integrated flux and the escape fraction with respect to the
rotational velocity.

@@

One way to understand this is that a rotating sphere
is identical from a static one. 
While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame.  
Therefore, in the frame of the rotating gas cloud all atoms are
stationary with respect to each other and the scattering process
proceeds identical as in the static case.

An alternative - more quantitative way - to view this is   to consider
the path of an individual photon: say the photon travels  a distance
corresponding to $\tau=1$, scatters by $90$ degrees, then travels
another distance corresponding to $\tau=1$. 
The point where the photon ends up is independent of the assumed
rotation velocity of the cloud.

The spectrum of \lya photons emerging from a rotating gas cloud is
also identical as for the static case in a frame that is co-rotating
with the cloud. 
However, the surface of cloud now moves in the lab-frame. 
Each surface-element on the rotating cloud now has a bulk
velocity with respect to a distant observer. 
In order to compute the
spectrum one can integrate over all the surface elements in the
sphere with their corresponding shift in velocity and an additional
weight by the surface intensity.

@@

The two methods clearly give good agreement, though not perfect. 
In particular, the left panel shows that the MC gives rise to a spectrum that is
slightly more concentrated towards the line centre. 
It is possible to get better agreement my modifying the surface brightness
profile.
In any case, the analytic calculation closely captures the results
obtained from the full calculations from the MC simulations.  
As such, they are extremely useful and provide us with a quick tool to
verify our calculations at the first order level.

@@

{\bf Escape at Line Center.} Our models have shown that rotation
enhances the flux density at line center (see Fig.     \ref{fig:differentvelocities}). It has
recently been proposed that galaxies with Lya spectral lines that
contain flux at line center may be `leaking' ionizing (LyC) photons
\citep{2014arXiv1404.2958V,Behrens2014}. The main reason for this possible
connection is that the escape of ionizing (LyC) photons requires
$N_{\rm HI} < 10^{17} $cm$^{-2}$. The same low column densities facilitate the escape of
\ly photons at (or close to) line center. Our work suggests that
rotation may provide an alternative explanation.

@@

The definitive and clear impact of rotation on the \ly morphology
suggests that this is an effect that should be taken into account at
the moment of interpreting high resolution spectroscopic data. In
particular it is relevant to consider the joint effect of rotation the
and ubiquitous outflows (Remolina-Gutierrez et al., in prep.)
because rotation can lead to enhanced escape of \ly at line center, which
has also been associated with escape of ionizing (LyC) photons
\cite{Behrens2014,2014arXiv1404.2958V}

@@

\appendix
\section{Analytic Expression for the Ly$\alpha$ Spectrum
  emerging from Rotating Cloud}

>>>>

A basic understanding of the quantitative behavior of the \ly line
has been reached through analytic studies in the case of a static
configurations, such as uniform slabs
\citep[][]{Adams72,Harrington73,Neufeld90} and uniform spheres
\citep{Dijkstra06}. 
Analytic studies of configurations including some kind of bulk flow
only include the case of a sphere with a Hubble like expansion flow
\citep{LoebRybicki}.

A more detailed quantitative description of the \ly line has been
reached through Monte Carlo simulations \citep{Auer68,Avery68,Adams72}. 
In the last two decades these studies have become popular due to the
availability of computing power. 
Early into the 21st century, the first
studies focused on homogeneous and static media
\citep{Ahn00,Ahn01,Zheng02}. 
Later on, the effects of clumpy media \citep{Hansen06} and
expanding/contracting shell/spherical geometries started to be studied
\citep{Ahn03,Verhamme06,Dijkstra06}.  For a recent review, we refer the interested reader to \citet{review}.
Similar codes have applied these results to semi-analytic models of
galaxy formation \citep{Orsi12, Garel2012} and results of large
hydrodynamic simulations \citep{CLARA,Forero12,Behrens13}. 
Recently, Monte Carlo codes have also been applied to the results of
high resolution hydrodynamic simulations of individual galaxies
\citep{Laursen09,Barnes11,Verhamme12,Yajima12}. 
Meanwhile, recent  developments have been focused on the systematic
study of clumpy outflows \citep{DijkstraKramer} and anisotropic
velocity configurations \citep{Zheng2013}.

@@

\subsection{Integrated Line Intensity}
\label{sec:intlineint}

@@

Fig. \ref{fig:muhisto} shows the results for a selection of models
with $\tau_{\rm H}=10^{5}$, different rotational velocities and the two
types of source distributions. This shows that $F(\mu)$ is consistent with being flat, apart
from some statistical fluctuations on the order of 2\%.

This is a remarkable result: while the rotation axis defines preferential direction, the
integrated flux is the same for all viewing angles in the range of parameters explored in this paper. {\bf This can be understood from the fact that
{\it radiative transfer inside a sphere that undergoes solid-body rotation proceeds identical as inside a static sphere}: we can draw a line between any two atoms within the rotating cloud, and their relative velocity along this line is zero (apart from the relative velocity as a result of random thermal motion), irrespective of the rotation velocity of the cloud. This relative velocity is what is relevant for the radiative transfer\footnote{This point can be further illustrated by considering the path of individual photons: let a photon be emitted at line center ($x=0$), in some random direction ${\bf k}$, propagate a distance that corresponds to $\tau_0=1$, scatter fully coherently (i.e. $x=0$ after scattering in the gas frame) by 90$^{\circ}$, and again propagate a distance that corresponds to $\tau_0=1$. The position where the photon scatters next does {\it not} depend on the rotation of the cloud, nor on ${\bf k}$.}.}

@@

We now estimate the escape fraction $f_{\rm esc}$ for the dusty
models. The main result is that we do not find any significant dependence
with either the viewing angle nor the rotational velocity. This is consistent with our finding in \S~\ref{sec:intlineint}, that radiative transfer inside the cloud does not depend on its rotational velocity. For completeness we list in Table \ref{table:escape} the escape
fraction for all models.

@@

The number of scatterings affects the escape frequency of a \ly
photon. Studying this quantity further illustrates the independence of
the integrated flux and the escape fraction on rotational velocity.

@@

As we explained in \S~\ref{sec:intlineint}, the best way to understand this is that radiative transfer inside a sphere that undergoes solid-body rotation
proceeds identical to that inside a non-rotating sphere. While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame. Therefore, in the frame of the rotating gas cloud all atoms are
stationary with respect to each other and the scattering process
proceeds identical as in the static case (also see \S~\ref{sec:intlineint} for an additional more quantitative explanation).

%An alternative - more quantitative way - to view this is   to consider the path of an individual photon: say the photon travels  a distance corresponding to $\tau=1$, scatters by $90$ degrees, then travels another distance corresponding to $\tau=1$. The point where the photon ends up is independent of the assumed rotation velocity of the cloud.

This result allows us to analytically estimate the spectrum emerging from a rotating cloud:
The spectrum of \lya photons emerging from a rotating gas cloud is identical as for the static case in a frame that is co-rotating
with the cloud. However, the surface of cloud now moves in the lab-frame. 
Each surface-element on the rotating cloud now has a bulk
velocity with respect to a distant observer. In order to compute the
spectrum one can integrate over all the surface elements in the
sphere with their corresponding shift in velocity and an additional
weight by the surface intensity.

@@

The two methods clearly give good agreement, though not perfect.  In particular, the left panel shows that the MC gives rise to a spectrum that is
slightly more concentrated towards the line centre. As we explain in Appendix~\ref{sec:app}, we do not expect perfect agreement, because this requires an analytic solution for the spectrum of Ly$\alpha$ photons emerging from a static, optically extremely thick cloud {\it as a function of the angle at which they escape from the sphere}. This solution does not exist in the literature. It is possible to get better agreement my modifying the surface brightness profile.
In any case, the analytic calculation closely captures the results
obtained from the full calculations from the MC simulations.  
As such, they are extremely useful and provide us with a quick tool to
verify our calculations at the first order level.

@@

{\bf Escape at Line Center.} Our models have shown that rotation
enhances the flux density at line center (see Fig.     \ref{fig:differentvelocities}). It has
recently been proposed that galaxies with Lya spectral lines that
contain flux at line center may be `leaking' ionizing (LyC) photons
\citep{Behrens2014,2014arXiv1404.2958V}. The main reason for this possible
connection is that the escape of ionizing (LyC) photons requires
$N_{\rm HI} < 10^{17} $cm$^{-2}$. The same low column densities facilitate the escape of
\ly photons at (or close to) line center. Our work suggests that
rotation may provide an alternative explanation.

@@

The definitive and clear impact of rotation on the \ly morphology
suggests that this is an effect that should be taken into account at
the moment of interpreting high resolution spectroscopic data. In
particular it is relevant to consider the joint effect of rotation the
and ubiquitous outflows (Remolina-Gutierrez et al., in prep.)
because rotation can lead to enhanced escape of \ly at line center, which
has also been associated with escape of ionizing (LyC) photons
\citep{Behrens2014,2014arXiv1404.2958V}

@@

\appendix
\section{Analytic Expression for the Ly$\alpha$ Spectrum
  emerging from Rotating Cloud}  
\label{sec:app}
